---
title: "Reproducible Research Course Project 1"
author: "Kalli Buchanan"
date: "24/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

### Loading and Preprocessing the Data

Download data from provided url and load as dataframe.
```{r loading, results='hide'}
library(readr)

if(!file.exists(".data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "./data/activity_monitoring.csv", method = "curl")
activityData <- read_csv("data/activity_monitoring.csv")

```

Data is already processed, so no need to process further.
```{r dataset_head}
head(activityData)
```

Let's take a quick look at what the data is comprised of.
```{r summary}
summary(activityData)
```

### Mean total number of steps taken

Calculate the total number of steps taken per day and create new data.frame to 
hold this information
```{r total_steps}
stepsData <- with(activityData, data.frame(date = unique(date), 
                                           total.steps = tapply(steps, date, 
                                                                sum, na.rm = TRUE)))
head(stepsData)
```
Histogram of the total amount of steps by each day.  
```{r hist_steps}
library(ggplot2)
g <- ggplot(stepsData, aes(total.steps))
g + geom_histogram(colour = "white") +
  geom_vline(aes(xintercept=mean(total.steps)), colour="blue", lty="dashed", size=0.5) +
  labs(title = "Histogram of Total Steps by Date",
       x = "Total Steps",
       y = "Count")
```  

  
Find the mean and median of total number of steps taken per day.
```{r mean_median}
stepsMean <- round(mean(stepsData$total.steps), digits = 2)
stepsMedian <- round(median(stepsData$total.steps), digits = 2)
```
The mean is **`r stepsMean`** and the median is **`r stepsMedian`**.

### Average Daily Activity Pattern

Make a new data.frame which holds unique intervals with the number of steps 
averaged across all days.

```{r intervals}
intervalData <- with(activityData, data.frame(interval = unique(interval), 
                                              average.steps = tapply(steps, interval, 
                                                                     mean, na.rm = TRUE)))
head(intervalData)
```
View these results using a time series plot.  Dashed vertical line shows the 
interval with the highest average steps.

```{r timeseries}
ts <- ggplot(intervalData, aes(interval, average.steps))
ts + geom_line() + 
  geom_vline(aes(xintercept = interval[average.steps == max(average.steps)]), col = "blue", lty = "dashed") +
  labs(title = "Daily Average Steps by Five Minute Interval",
       x = "Five Minute Interval (spans one day)",
       y = "Average Steps (averaged across all days)")
```  
  
Find the interval with the highest average steps across all days.
```{r max_interval}
maxInterval <- with(intervalData, interval[average.steps == max(average.steps)])
maxAvgSteps <- with(intervalData, round(average.steps[interval == maxInterval], digits = 2))
```
The interval **`r maxInterval`**, has the highest average steps (**`r maxAvgSteps`**).

### Impute Missing Values

Find number of missing values in the data set.

```{r missing_values}
sum(is.na(activityData))
```

Fill in missing values in data set using an average of the previous and next day 
steps for that specific interval.

```{r imputation}
imputedData = activityData

for (i in 1:dim(imputedData)[1]) {
  if (is.na(imputedData$steps[i])){
    before <- ifelse(imputedData$date[i] == "2012-10-01", NA, 
                     imputedData$steps[imputedData$date == imputedData$date[i]-1 & 
                                         imputedData$interval == imputedData$interval[i]])
    after <- ifelse(imputedData$date[i] == "2012-11-30", NA, 
                    imputedData$steps[imputedData$date == imputedData$date[i]+1 
                                      & imputedData$interval == imputedData$interval[i]])
    newSteps <- mean(c(before,after), na.rm = TRUE)
    imputedData$steps[i] = newSteps
  }
}
sum(is.na(imputedData))

```
This shows that all missing values have been imputed.  
  
Re-vist finding from original calculations when missing values were ignored.  
  
First create a new data frame with total steps per day.
```{r imputed_total_steps}
imputedSteps <- with(imputedData, data.frame(date = unique(date), 
                                           total.steps = tapply(steps, date, 
                                                                sum, na.rm = FALSE)))
head(imputedSteps)
```
Histogram of the total steps per day.
```{r imputed_hist}
ig <- ggplot(imputedSteps, aes(total.steps))
ig + geom_histogram(colour = "white") +
  geom_vline(aes(xintercept=mean(total.steps)), colour="blue", lty="dashed", size=0.5) +
  labs(title = "Histogram of Total Steps by Date",
       x = "Total Steps",
       y = "Count")
```

  
Find the mean and median of total number of steps taken per day.
```{r imputed_mean_median}
imputedMean <- round(mean(imputedSteps$total.steps), digits = 2)
imputedMedian <- round(median(imputedSteps$total.steps), digits = 2)
```
The mean is **`r imputedMean`** and the median is **`r imputedMedian`**.  These 
are both larger than the original values by **`r imputedMean - stepsMean`** and 
**`r imputedMedian - stepsMedian`**, respectively.  

### Weekdays and Weekends Activity Patterns

Add new factor variable to data set to indicate whether the date is during the 
week or weekend.
```{r add_factor}
library(dplyr)
weekendday <- c("Saturday", "Sunday")
weekData <- imputedData %>%
  mutate(date.type = ifelse(weekdays(date) %in% weekendday, "weekend", "weekday"), date.type = as.factor(date.type))
summary(weekData)
```
Make two new data.frame which holds unique intervals with the number of steps 
averaged across all days; one for weekdays, and one for weekends. Join these 
tables based on interval.

```{r intervals_weekday}
library(tidyr)
weekdayData <- with(weekData[weekData$date.type == "weekday",], data.frame(interval = unique(interval), 
                                              average.steps.weekday = tapply(steps, interval, 
                                                                     mean)))
weekendData <- with(weekData[weekData$date.type == "weekend",], data.frame(interval = unique(interval), 
                                              average.steps.weekend = tapply(steps, interval, 
                                                                     mean)))
weekInterval <- merge(weekdayData, weekendData) %>%
  pivot_longer(!interval, names_to = "date.type", values_to = "average.steps", names_prefix = "average.steps.") %>%
  mutate(date.type = as.factor(date.type))

head(weekInterval)
```

View these results using a time series plot.

```{r timeseries_weekday}
wd <- ggplot(weekInterval, aes(interval, average.steps))
wd + facet_grid(date.type~.) +
  geom_line() + 
  labs(title = "Daily Average Steps by Five Minute Interval",
       x = "Five Minute Interval (spans one day)",
       y = "Average Steps (averaged across all days)")

```  