---
title: "Reproducible Research Course Project 1"
author: "Kalli Buchanan"
date: "24/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

### Loading and Preprocessing the Data

Download data from provided url and load as dataframe.
```{r loading, results='hide'}
library(readr)

if(!file.exists(".data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "./data/activity_monitoring.csv", method = "curl")
activityData <- read_csv("data/activity_monitoring.csv")

```

Data is already processed, so no need to process further.
```{r dataset_head}
head(activityData)
```

Let's take a quick look at what the data is comprised of.
```{r summary}
summary(activityData)
```

### Mean total number of steps taken

Calculate the total number of steps taken per day and create new data.frame to 
hold this information
```{r total_steps}
stepsData <- with(activityData, data.frame(date = unique(date), 
                                           total.steps = tapply(steps, date, 
                                                                sum, na.rm = TRUE)))
head(stepsData)
```
Histogram of the total amount of steps by each day.  
```{r hist_steps}
library(ggplot2)
g <- ggplot(stepsData, aes(total.steps))
g + geom_histogram(colour = "white") +
  geom_vline(aes(xintercept=mean(total.steps)), colour="blue", lty="dashed", size=0.5) +
  labs(title = "Histogram of Total Steps by Date",
       x = "Total Steps",
       y = "Count")
```  

  
Find the mean and median of total number of steps taken per day.
```{r mean_median}
stepsMean <- round(mean(stepsData$total.steps), digits = 2)
stepsMedian <- round(median(stepsData$total.steps), digits = 2)
```
The mean is **`r stepsMean`** and the median is **`r stepsMedian`**.

### Average Daily Activity Pattern

Make a new data.frame which holds unique intervals with the number of steps 
averaged across all days.

```{r intervals}
intervalData <- with(activityData, data.frame(interval = unique(interval), 
                                              average.steps = tapply(steps, interval, 
                                                                     mean, na.rm = TRUE)))
head(intervalData)
```
View these results using a time series plot.  Dashed vertical line shows the 
interval with the highest average steps.

```{r timeseries}
ts <- ggplot(intervalData, aes(interval, average.steps))
ts + geom_line() + 
  geom_vline(aes(xintercept = interval[average.steps == max(average.steps)]), col = "blue", lty = "dashed") +
  labs(title = "Daily Average Steps by Five Minute Interval",
       x = "Five Minute Interval (spans one day)",
       y = "Average Steps (averaged across all days)")
```  
  
Find the interval with the highest average steps across all days.
```{r max_interval}
maxInterval <- with(intervalData, interval[average.steps == max(average.steps)])
maxAvgSteps <- with(intervalData, round(average.steps[interval == maxInterval], digits = 2))
```
The interval **`r maxInterval`**, has the highest average steps (**`r maxAvgSteps`**).






